Simulating High-Dimensional Multivariate Data

### 2020-12-17 010-background.Rmd
4th draft. Address ADK feedback

* SP1: Procedure must scale to high dimensions, computable in a reasonable amount time.
* SP2: Procedure must scale to high dimensions while maintaining accuracy.

@@

*Discrete marginal considerations*.
Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties (for example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1; @MB13).
One remedy for this issue is to rescale Equation \@ref(eq:spearman). 
For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as

@@

*Marginal-dependent bivariate correlation bounds*. 
Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$.
The so-called *Frechet-Hoeffding bounds* are well-studied (see @Nelsen2007 and @BF17).
This gives strict restraints on the possible correlations and cannot be overcome through algorithm design.
In general, the bounds are given by

>>>>

* SP1: Procedure must scale to high dimensions with practical computable times.
* SP2: Procedure must scale to high dimensions while maintaining accuracy.

@@

*Discrete marginal considerations*.
Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties.
For example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1 [@MB13].
One remedy for this issue is to rescale Equation \@ref(eq:spearman). 
For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as

@@

*Marginal-dependent bivariate correlation bounds*. 
Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$.
The so-called *Frechet-Hoeffding bounds* are well-studied ([@Nelsen2007; @BF17]).
These constraints cannot be overcome through algorithm design.
In general, the bounds are given by

### 2020-12-17 020-simulation-algorithm.Rmd
4th draft. Address ADK feedback

This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,2,\ldots,d$.
Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$.
To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$.
The method is best understand as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.

@@

Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
Specifically, step 2i uses either an efficient multi-core multivariate normal simulator (the `R` package `mvnfast`[@Fasiolo2016]) or Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm from the preceding section).

>>>>

This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,2,\ldots,d$.
Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$.
To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$.
The method can be described as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.

@@

Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
Specifically, step 2i uses either the efficient multi-core multivariate normal simulator `mvnfast`[@Fasiolo2016] or Google's JAX python library for GPU acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm from the preceding section).

### 2020-12-17 030-bigsimr-package.Rmd
4th draft. Address ADK feedback

This subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result visualization.
For this low-dimensional setting, we operate using a single central processing unit (CPU), as the overhead in forking the tasks to multiple cores outweighs the computational gains. 
Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
The section concludes with a short description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package [@R-rslurm].

@@

It is a bit inconvenient to repeatedly fill in the parameter values manually, so we provide a convenience function `mlist` which behaves similarly to `alist`, except that it evaluates the right side of argument values within the list.
This is intended to help with scaling up your code to high dimensions when many marginals to specify.

@@

*Specifying correlation*.
As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
For non-normal data and for improved simulation accuracy in our scheme, we advocate the use of rank-based correlations Spearman's $\rho_S$ and Kendall's $\tau$.
But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Pearson, Spearman, or Kendall correlation using the fastest methods available.
(Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
These estimation methods are the standard approaches, not designed specifically designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).

@@

```{r ch030-rvecRslurmMap-comments, echo=FALSE, eval=FALSE}
library(rslurm)
## a single call of rvec
## had to edit rslurm/templates/submit_sh.txt
## source ~/.bashrc
## conda activate bigsimr
## conda env config vars list
set.seed(06202020)
simReps <- 100
rhoList <- replicate( n = simReps, bigsimr::cor_randPSD(d = length(margins)),
                     simplify=FALSE )
sjob <- slurm_map(x = rhoList,
                  f = rvec,
                  jobname = 'rvecMap',
                  n=1e6,
                  margins = margins,
                  type = "spearman",
                  nodes = 4,
                  cpus_per_node = 4,
                  submit = TRUE)

@@

```{r ch030-rvecRslurmMap, echo=TRUE, eval=FALSE}
library(rslurm)
simReps <- 100
rhoList <- replicate( n = simReps, bigsimr::cor_randPSD(d = length(margins)),
                     simplify=FALSE )
sjob <- slurm_map(x = rhoList,
                  f = rvec,
                  jobname = 'rvecMap',
                  n=1e6,
                  margins = margins,
                  type = "spearman",
                  nodes = 4,
                  cpus_per_node = 4,
                  cores = 4,
                  submit = TRUE)
```

>>>>

The subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the readily-available data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result visualization.
For this low-dimensional setting, we operate using a single central processing unit (CPU), as the overhead in forking the tasks to multiple cores outweighs the computational gains. 
Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
The section concludes with a short description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package [@R-rslurm].

@@

It is a bit inconvenient to repeatedly fill in the parameter values manually, so we provide a convenience function `mlist` which behaves similarly to `alist`, except that it evaluates the right side of argument values within the list.

@@

*Specifying correlation*.
As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
For non-normal data and for improved simulation accuracy in our scheme, we advocate the use of rank-based correlations Spearman's $\rho_S$ and Kendall's $\tau$.
But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Pearson, Spearman, or Kendall correlation using the fastest methods available.
(Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).

@@

```{r ch030-rvecRslurmMap-comments, echo=FALSE, eval=FALSE}
library(rslurm)
## a single call of rvec
## had to edit rslurm/templates/submit_sh.txt
## source ~/.bashrc
## conda activate bigsimr
## conda env config vars list
set.seed(06202020)
simReps <- 100
rhoList <- replicate(n = simReps, bigsimr::cor_randPSD(d = length(margins)),
                     simplify=FALSE)
sjob <- slurm_map(x = rhoList,
                  f = rvec,
                  jobname = 'rvecMap',
                  n=1e6,
                  margins = margins,
                  type = "spearman",
                  nodes = 4,
                  cpus_per_node = 4,
                  submit = TRUE)

@@

```{r ch030-rvecRslurmMap, echo=TRUE, eval=FALSE}
library(rslurm)
simReps <- 100
rhoList <- replicate(n = simReps, bigsimr::cor_randPSD(d = length(margins)),
                     simplify=FALSE)
sjob <- slurm_map(x = rhoList,
                  f = rvec,
                  jobname = 'rvecMap',
                  n=1e6,
                  margins = margins,
                  type = "spearman",
                  nodes = 4,
                  cpus_per_node = 4,
                  cores = 4,
                  submit = TRUE)
```

### 2020-12-17 040-monte-carlo-evaluation.Rmd
4th draft. Address ADK feedback

We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example and, so, proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples). 
We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's ---- in identical, bivariate marginal configurations.
For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds. 
The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).

>>>>

We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example and, so, proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples). 
We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's --- in identical, bivariate marginal configurations.
For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds. 
The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).

### 2020-12-17 050-examples.Rmd
4th draft. Address ADK feedback

This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions and with estimated parameters that mimic the observed data and its generating process.
We seek scaleable realistic multivariate simulation to enable Monte Carlo (MC) methods for these data.
Modeling RNA-seq using multivariate probabilities distributions is natural as inter-gene correlation is an inherent part of biological processes [@Wang2009b].
Yet many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
See @Efron2012 for a detailed discussion with related methods and @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.

@@

With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`:

```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE, cache=F}
## Estimate Spearman's correlation on the count data
corType <- 'spearman'
system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
```

@@

```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
## Set the number of random vectors
n <- 10000
## construct margins
nb_margins <- make_nbinom_alist(sizes, probs)
## run sims
sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
                   ensure_PSD = TRUE, cores = cores) 
colnames(sim_nbinom) <- names(brca)
```

@@

To conduct statistical inference a critical task is to evaluate the joint probability mass (or density) function:

@@

$$
\hat{P}( {\bf Y} >= {\bf y_0 } ) = \sum_{b=1}^B I( {\bf Y^{(b) }} > {\bf y_0 } ) / B
$$

```{r ch050-densityEvaluation, include=FALSE, eval=TRUE, cache=F}
d <- ncol(sim_nbinom)
B <- nrow(sim_nbinom)
threshold <- rep( 1, d)
pHat <- mean(apply( sim_nbinom,  1,
           function(X, y0=threshold) {
               all( X >= y0) }
           ))
```

>>>>

This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions and with estimated parameters that mimic the observed data and its generating process.
We seek realistic multivariate simulation to enable Monte Carlo (MC) methods for these data.
Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation is an inherent part of biological processes [@Wang2009b].
Yet many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
See @Efron2012 for a detailed discussion with related methods and @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.

@@

With this goal in mind, we first estimate the desired correlation matrix using the fast implementation provided by `bigsimr`.:

```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE, cache=F}
## Estimate Spearman's correlation on the count data
corType <- 'spearman'
nb_Rho <- bigsimr::cor_fast(brca, method = corType)
```

@@

```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
## Set the number of random vectors (B in text)
n <- 10000
## construct margins
nb_margins <- make_nbinom_alist(sizes, probs)
## run sims
sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
                   ensure_PSD = TRUE, cores = cores) 
colnames(sim_nbinom) <- names(brca)
```

@@

Many statistical tasks require evaluation of a joint probability mass (or density) function:

@@

$$
\hat{P}( {\bf Y} >= {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y^{(b) }} >= {\bf y_0 } )
$$

```{r ch050-densityEvaluation, include=FALSE, eval=TRUE, cache=F}
d <- ncol(sim_nbinom)
B <- nrow(sim_nbinom)
threshold <- rep( 1, d)
pHat <- mean(apply(sim_nbinom,  1,
           function(X, y0=threshold) {
               all(X >= y0)}
           ))
```

### 2020-12-17 060-conclusion-discussion.Rmd
4th draft. Address ADK feedback

It is customary to compare new tools and algorithms directly to existing competing methods and software.
In this study, however, we only employ our proposed methodology, as our previous work has show that existing tools are simply not designed nor feasible to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula`package and others). 
For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.

>>>>

It is customary to compare new tools and algorithms directly to existing competing methods and software.
In this study, however, we only employ our proposed methodology, as our previous work has shown that existing tools are simply not designed nor feasible to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula` package and others). 
For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.

### 2020-12-17 index.Rmd
4th draft. Address ADK feedback

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and evaluating quantitative methods. But measurements are often correlated and high dimensional in this era of big data, however, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$.
---

@@

With all this mind, we present a scaleable, flexible multivariate simulation algorithm. 
The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
Further, we introduce the `bigsimr` R package that provides parallelized, high-performance software implemented in our NORTA-inspired algorithm.
The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependency measures for the multivariate normal (MVN) distribution.

Our study proceeds by providing background information, including a description of our motivating example application --- RNA-sequencing (RNA-seq) breast cancer data.
Then we describe and justify our simulation methodology and related algorithms.
Next, we detail an illustrative low-dimensional example of basic and advanced use of the `bigsimr` R package.
Then we proceed with Monte Carlo studies under various distributional assumptions to assess accuracy and scaleable.
After the MC evaluations, we revisit our high-dimensional motivating example and employ our methods in commonplace statistical computing tasks.
Finally, we discuss the method's utility, limitations, and future directions.

>>>>

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and evaluating quantitative methods. But measurements are often correlated and high dimensional in this era of big data, however, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$.
---

@@

With all this mind, we present a scaleable, flexible multivariate simulation algorithm. 
The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
Further, we introduce the `bigsimr` R package that provides high-performance software implementing our algorithm.
The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependency measures for the multivariate normal (MVN) distribution.

Our study proceeds by providing background information, including a description of our motivating example application --- RNA-sequencing (RNA-seq) breast cancer data.
Then we describe and justify our simulation methodology and related algorithms.
Next, we detail an illustrative low-dimensional example of basic and advanced use of the `bigsimr` R package.
Then we proceed with Monte Carlo studies under various distributional assumptions to assess accuracy and computation times for increasingly large $d$.
After the MC evaluations, we revisit our high-dimensional motivating example and employ our methods in commonplace statistical computing tasks.
Finally, we discuss the method's utility, limitations, and future directions.
